What Can Stack Traces Tell us about the Use of Exceptions in Java?

### 2014-05-09 android-stacks.tex
small corrections 1

In this work, we perform the first large scale analysis of Java stack traces and
pinpoint how they can reveal bad programming practices. We mined and analyzed
the stack traces embedded in all issues of Java projects available GitHub.
Overall, our research set includes 356,057 issues define at 16,836 projects
(from which 28,800 stack traces were extracted). From this set, the stack traces
of 482 Android projects were investigated in more detail, in combination with
source code and bytecode analysis. In this study some patterns of exception
(mis)-use were consistently detected such as: unexpected wrappings (e.g., Errors
being wrapped in checked exceptions), revealing that Java hybrid exception
model is not fully used according to its purpose;  undocumented runtime
exceptions signaled by third party code, which makes almost impossible for
library clients to protect against such exceptions; and a high prevalence of
java.lang exceptions reported on issues, representing approximately 50\% of the
analyzed issues.

@@

Often some languages constructions may lead the developers into believing that
by just re-throwing the exceptions they can forget about the exceptional
situations during the development of the "happy path". This "ignore-for-now"
approach may turn the exception handling into a generalized "goto"
mechanism~\cite{mandrioli1992advances} making the program more complex and even
less reliable. This behaviour may lead to the well known uncaught exceptions
problem~\cite{jo2004uncaught}, one of the main causes of application crashes.

@@

This analysis also revealed unexpected wrappings such as: Error wrapping a Checked 
exception; a Runtime wrapping and Error; and an Error wrapping a Runtime.
Java is the only language that provides a hybrid exception handling mechanism
which offers  different types to represent different exception behaviors (i.e., error, 
runtime and checked) (see Section ~\ref{extypes}) . According to Java specification Errors should not be
handled inside the system since they usually represent unrecoverable conditions
detected by the JVM such as OutOfMemoryError. Checked exceptions should
represent recoverable conditions and  Runtime exceptions represent are usually
related to programming mistakes from which the developer is not expected to recover
from.

@@

Since there is no way of enforcing Java exception type conventions during program development,
we could observe the stack trace analysis combined with types these three types
of exceptions were used interchangeably. This interchangeability makes the
behaviour of the exception handling code more complex and less reliable. For
instance, considering a situation were an OutOfMemoryError (a situation that
should not be handled) was wrapped in a checked exception (a situation that must
he handled), in this case this wrapping is telling to the caller that it should
handled.

@@

In this work we present an exploratory study on which we mined the stack 
traces embedded in all issues of Java projects available GitHub, and analyzed the stack traces
reported on a subset of 482 Android projects in more detail. In this study the information extracted 
from stack traces was used in combination with source code and bytecode analysis to 
pin-point patterns of exceptions (mis-)use in Java. And some patterns of exception
(mis)-use were consistently detected such as: unexpected wrappings (e.g., Errors
being wrapped in checked exceptions), revealing that Java hybrid exception
model is not fully used according to its purpose; undocumented runtime
exceptions signaled by third party code (which makes almost impossible for
library clients to protect against such exceptions); and a high prevalence of
java.lang exceptions reported on issues (representing approximately 50\% of the
analyzed issues). Such (mis)uses detected in out study can negatively affect the 
robustness of current Java software systems. Such results shed light to needs for  
 developing tools support to help developers dealing with such problems and 
improving the design of exception handling mechanism in Java.

>>>>

In this work, we perform the first large scale analysis of Java stack traces and
pinpoint how they can reveal bad programming practices. We mined and analyzed
the stack traces embedded in all issues of Java projects available GitHub.
Overall, our research set includes 356,057 issues defined at 16,836 projects, 
from which 28,800 stack traces were extracted. From this set, the stack traces
of 482 Android projects were investigated in more detail, in combination with
source code and bytecode analysis. In this study some patterns of exception
(mis)-use were consistently detected such as: unexpected wrappings (e.g., Errors
being wrapped in checked exceptions), revealing that Java hybrid exception
model is not fully used according to its purpose;  undocumented runtime
exceptions signaled by third party code, which is a serious threat to program
dependability; and a high prevalence of java.lang exceptions reported on issues,  
causing approximately 50\% of the analyzed stack traces.

@@

Often some exception handling constructions may lead the developers into believing that
by just re-throwing the exceptions they can forget about the exceptional
situations during the development of the "happy path". This "ignore-for-now"
approach may turn the exception handling into a generalized "goto"
mechanism~\cite{mandrioli1992advances} making the program more complex and even
less reliable. This behaviour may lead to the well known uncaught exceptions
problem~\cite{jo2004uncaught}, one of the main causes of application crashes.

@@

This analysis also revealed unexpected wrappings such as: Error wrapping a Checked 
exception; a Runtime wrapping and Error; and an Error wrapping a Runtime.
Java is the only language that provides a hybrid exception handling mechanism
which offers  different types to represent different exception behaviors (i.e., error, 
runtime and checked) (see Section ~\ref{sec:best}) . According to Java specification Errors should not be
handled inside the system since they usually represent unrecoverable conditions
detected by the JVM such as OutOfMemoryError. Checked exceptions should
represent recoverable conditions and  Runtime exceptions represent are usually
related to programming mistakes from which the developer is not expected to recover
from.

@@

Since there is no way of enforcing Java exception type conventions during program development,
we could observe the stack trace analysis combined with types these three types
of exceptions were used interchangeably. This interchangeability makes the
behaviour of the exception handling code more complex and less reliable. For
instance, considering a situation were an instance of OutOfMemoryError (a situation that
should not be handled) was wrapped in a checked exception (a situation that must
he handled), in this case this wrapping is telling to the caller that it should
handled.

@@

In this work we present an exploratory study on which we mined the stack 
traces embedded in all issues of Java projects available GitHub, and analyzed in mode detail thr
stack traces reported on a subset of 482 Android projects. In this study, the information extracted 
from stack traces was used in combination with source code and bytecode analysis to 
pin-point patterns of exceptions (mis-)use in Java. And some patterns of exception
(mis)-use were consistently detected such as: unexpected wrappings (e.g., Errors
being wrapped in checked exceptions), revealing that Java hybrid exception
model is not fully used according to its purpose; undocumented runtime
exceptions signaled by third party code (which makes almost impossible for
library clients to protect against such exceptions); and a high prevalence of
java.lang exceptions reported on issues (representing approximately 50\% of the
analyzed issues). Such (mis)uses detected in out study can negatively affect the 
robustness of current Java software systems. Such results shed light to needs for  
 developing tools support to help developers dealing with such problems and 
improving the design of exception handling mechanism in Java.
